---
title: "Take_home_assignment"
author: "Vicky (Ke) Xu"
date: "15/02/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```


## City of Toronto
## PARKS, FORESTRY & RECREATION DIVISION
## Data Analyst & Integrator (Data Science)
## Job ID# 13025(X)
## Technical Assignment Result

### Load all requried packages
```{r echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidygeocoder)
library(jsonlite)
library(sp)
library(sf)
library(geosphere)
library(leaflet)
library(lubridate)
```

### The very first step is to load all parking ticket data. I took it from year 2016 to 2018.
```{r}
all_file_names <- list.files('parking-ticket/parking-tickets-2016/',full.name=TRUE)
listDFs <- lapply(all_file_names, read.csv, quote='', stringsAsFactors=FALSE)
combDFs_2016 <- do.call("rbind", listDFs)

all_file_names <- list.files('parking-ticket/parking-tickets-2017/',full.name=TRUE)
listDFs <- lapply(all_file_names, read.csv, quote='', stringsAsFactors=FALSE)
combDFs_2017 <- do.call("rbind", listDFs)

all_file_names <- list.files('parking-ticket/parking-tickets-2018/',full.name=TRUE)
listDFs <- lapply(all_file_names, read.csv, quote='', stringsAsFactors=FALSE)
combDFs_2018 <- do.call("rbind", listDFs)


parking_ticket <- rbind(combDFs_2016,combDFs_2017)
parking_ticket <- rbind(parking_ticket,combDFs_2018)
```

### Preliminary examinations on missing values
```{r}
# Check the number of missing values within each column

parking_ticket_missing_summary <- apply(parking_ticket,2,function(x) sum(!complete.cases(x))) 
print(parking_ticket_missing_summary)

# Remove missing values and $0 set_fine_amount (there are 93 of them)
parking_ticket <- parking_ticket[complete.cases(parking_ticket$date_of_infraction),]
parking_ticket <- parking_ticket[complete.cases(parking_ticket$infraction_code),]
parking_ticket <- parking_ticket[complete.cases(parking_ticket$set_fine_amount),]
parking_ticket <- parking_ticket[parking_ticket$set_fine_amount != 0,]

# Decided to leave all missing time_of_infraction as is because 
# it is not within the objectives of this assignment.
nrow(parking_ticket[!complete.cases(parking_ticket$time_of_infraction),])
```


### 3.1.1 Top 20 ticket infractions (frequency)
```{r}
top_20_freq <- parking_ticket %>%
  group_by(infraction_code,infraction_description) %>%
  summarise(freq=n()) %>%
  arrange(desc(freq)) %>%
  head(20) %>%
  as.data.frame()
top_20_freq
```

### 3.1.2 Top 20 ticket infractions (revenue)
```{r}
top_20_revenue <- parking_ticket %>%
  group_by(infraction_code,infraction_description) %>%
  summarise(revenue=sum(set_fine_amount)) %>%
  arrange(desc(revenue)) %>%
  head(20) %>%
  as.data.frame()
top_20_revenue
```


### 3.1.4a How far (as the crow flies) is the closest parking lot (Green P) to each of the top 20 infraction locations?

```{r}
top_20_locations <- parking_ticket %>%
  group_by(location2) %>%
  summarise(freq=n()) %>%
  arrange(desc(freq)) %>%
  head(20) %>%
  as.data.frame()

# Paste all information to make it ready for geocoding
top_20_locations$location_new <- paste0(top_20_locations$location2,',TORONTO,ON,CANADA')

# Geocode all 20 addresses using the geocode function from tidygeocoder with method='osm'
top_20_locations_geocoded <- top_20_locations %>% 
  tidygeocoder::geocode(location_new,method='osm')

# Load green-p-parking data
green_p <- fromJSON('green-p-parking/green-p-parking-2019.json') %>% as.data.frame()

# Convert carparks.lat and carparks.lng to numeric
green_p$carparks.lat <- as.numeric(green_p$carparks.lat)
green_p$carparks.lng <- as.numeric(green_p$carparks.lng)

# Make sure that there is no missing locations
nrow(green_p[!complete.cases(green_p$carparks.lat),])
nrow(green_p[!complete.cases(green_p$carparks.lng),])

# Create spatial objects for top_20_locations_geocoded and green_p
sp_top_20_locations_geocoded <- top_20_locations_geocoded
coordinates(sp_top_20_locations_geocoded) <- ~long+lat

sp_green_p <- green_p
coordinates(sp_green_p) <- ~carparks.lng+carparks.lat

# Create a new data frame with two addtional columns to store results
top_20_locations_w_min_parking_dist <- top_20_locations_geocoded
top_20_locations_w_min_parking_dist['min_dist_ind'] <- 0
top_20_locations_w_min_parking_dist['min_dist'] <- 0

# Loop through all 20 locaitons and find the closest parking location
for (i in 1:20)
{
  dist_mat <- distm(coordinates(sp_top_20_locations_geocoded[i,]),coordinates(sp_green_p))
  top_20_locations_w_min_parking_dist$min_dist_ind[i] <- which.min(dist_mat)
  top_20_locations_w_min_parking_dist$min_dist[i]  <- min(dist_mat)
}

green_p_new <- green_p %>%
  mutate(index=1:nrow(green_p))

top_20_locations_w_min_parking_dist<- left_join(top_20_locations_w_min_parking_dist, 
                                                green_p_new,by=c('min_dist_ind' = 'index')) %>%
                                                select(-min_dist_ind)
                                                as.data.frame()
```


### 3.1.4b How far (as the crow flies) is the closest TTC stop to the top 20 infraction locations?

```{r}
# Load TTC data
ttc_stops <- read.table('TTC Routes and Schedules/opendata_ttc_schedules/stops.txt',
                        header=TRUE,
                        sep=',',
                        stringsAsFactors = FALSE,
                        quote = "")
head(ttc_stops)
dim(ttc_stops)

ttc_stops$stop_lat <- as.numeric(ttc_stops$stop_lat)
ttc_stops$stop_lon <- as.numeric(ttc_stops$stop_lon)

sp_ttc_stops <- ttc_stops
coordinates(sp_ttc_stops) <- ~stop_lon+stop_lat

# Create two columns to store results
top_20_locations_w_min_ttc_stops <- top_20_locations_geocoded
top_20_locations_w_min_ttc_stops['min_dist_ind'] <- 0
top_20_locations_w_min_ttc_stops['min_dist'] <- 0

# Loop through all 20 locations and find the closest parking location
for (i in 1:20)
{
  dist_mat <- distm(coordinates(sp_top_20_locations_geocoded[i,]),coordinates(sp_ttc_stops))
  top_20_locations_w_min_ttc_stops$min_dist_ind[i] <- which.min(dist_mat)
  top_20_locations_w_min_ttc_stops$min_dist[i]  <- min(dist_mat)
}

ttc_stops_new <- ttc_stops %>%
  mutate(index=1:nrow(ttc_stops))

top_20_locations_w_min_ttc_stops<- left_join(top_20_locations_w_min_ttc_stops, 
                                             ttc_stops_new,by=c('min_dist_ind' = 'index')) %>%
                                             select(-min_dist_ind) %>%
                                             as.data.frame()
top_20_locations_w_min_ttc_stops
```


### 3.1.5a Impact of day of week in all infractions

```{r}
# First create columns for weekday, month, season and year

parking_ticket$month <- as.factor(month(ymd(parking_ticket$date_of_infraction)))
parking_ticket$year <- as.factor(year(ymd(parking_ticket$date_of_infraction)))
parking_ticket$weekday <- weekdays(as.Date(parking_ticket$date_of_infraction,
                 tryFormats = c('%Y%m%d')))

getSeason <- function(input.date){
  numeric.date <- 100*month(input.date)+day(input.date)
  ## input Seasons upper limits in the form MMDD in the "break =" option:
  cuts <- base::cut(numeric.date, breaks = c(0,319,0620,0921,1220,1231)) 
  # rename the resulting groups (could've been done within cut(...levels=) if "Winter" wasn't double
  levels(cuts) <- c("Winter","Spring","Summer","Fall","Winter")
  return(cuts)
}
parking_ticket$season <- getSeason(ymd(parking_ticket$date_of_infraction))

parking_ticket %>% 
  group_by(weekday) %>%
  summarise(number_of_infractions = n())
```


### 3.1.5b Impact of month of week in all infractions

```{r}
parking_ticket %>% 
  group_by(month) %>%
  summarise(number_of_infractions = n())
```


### 3.1.5c Impact of season of week in all infractions

```{r}
parking_ticket %>% 
  group_by(season) %>%
  summarise(number_of_infractions = n())
```

### 3.3.1.a.i Distribution of infractions by year

```{r}
parking_ticket %>%
  group_by(year) %>%
  summarise(count=n()) %>%
  ggplot(aes(x=year,y=count)) +   
    geom_bar(aes(fill = year), position = "dodge", stat="identity")+
    # geom_point(aes(x = year, y = count), size = 1.5, color="black", group = 2) +
    # geom_line(aes(x = year, y = count), size = 1.5, color="red", group = 1) +
    ylim(0,3000000) +
    theme(legend.position = "none")
```


### 3.3.1.a.ii Distribution of infractions by month

```{r}
parking_ticket %>%
  group_by(month) %>%
  summarise(count=n()) %>%
  ggplot(aes(x=month,y=count)) +   
  geom_bar(aes(fill = month), position = "dodge", stat="identity")+
  ylim(0,700000)+
  theme(legend.position = "none")
  # geom_point(aes(x = month, y = count), size = 1.5, color="black", group = 1) +
  # geom_line(aes(x = month, y = count), size = 1.5, color="blue", group = 1)
```


### 3.3.1.b Distribution of top 20 infractions by fines

```{r}
top_20_revenue <- top_20_revenue[order(top_20_revenue$infraction_code),]
top_20_revenue$infraction_code <- as.factor(top_20_revenue$infraction_code)
top_20_revenue %>%
  ggplot(aes(x=infraction_code,y=revenue)) +   
  geom_bar(aes(fill = infraction_code), position = "dodge", stat="identity")+
  theme(legend.position = "none")
```


### 3.3.2.a Geographic distribution (location) of top 20 infractions (count)

```{r}
top_20_infraction_location_map <- leaflet(sp_top_20_locations_geocoded) %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = -79.372573, lat = 43.679434, zoom = 10) %>%
  addMarkers(popup = paste("Address: ", sp_top_20_locations_geocoded$location2, "<br>",
                           "Total number of infractions: ", sp_top_20_locations_geocoded$freq, "<br>"))
top_20_infraction_location_map
```


### 3.3.2.b Geographic distribution by ward for top 20 infractions (count)
### Interpretation: find the total number of infractions within each ward for all top 20 infraction locations.

```{r}
# Import City of Toronto ward

sf_city_ward <- sf::st_read('City Wards/City Wards Data/City Wards Data.shp', 
                            stringsAsFactors=F, options = "ENCODING=UTF8")

city_ward_col_names <- read.csv('City Wards/City Wards Data/City Wards Data_fields.csv',
                                stringsAsFactors = F)
colnames(sf_city_ward) <- city_ward_col_names$name
head(sf_city_ward)

# First determine the polygon each location belongs to

top_20_locations_geocoded$AREA_NAME <- apply(top_20_locations_geocoded[,c('long','lat')], 1, function(row) {  
  # Transformation to palnar (epsg:2163) is required, since sf library assumes planar projection 
  sf_city_ward_pl <- st_transform(sf_city_ward$geometry, 2163)
  coords <- c(as.numeric(row[1]),as.numeric(row[2]))
  pnt_sf <- st_transform(st_sfc(st_point(coords),crs = 4326), 2163)
  # Use st_intersects to see which polygon does the point intersects with
  area <- sf_city_ward[which(st_intersects(pnt_sf, sf_city_ward_pl, sparse = FALSE)), ]$AREA_NAME 
  return(area)
})

top_20_locations_geocoded

# Then create a frequency table

count_20_locations_by_ward <- top_20_locations_geocoded %>%
  group_by(AREA_NAME) %>%
  summarise(count=sum(freq)) %>%
  select(AREA_NAME,count) %>%
  right_join(sf_city_ward,by='AREA_NAME') %>%
  select(AREA_NAME,count,geometry) %>%
  replace(is.na(.), 0)

# Finally, plot it out

unique_color_df <- data.frame(count = sort(unique(count_20_locations_by_ward$count)),
                              color = gsub(".{2}$","",topo.colors(length(unique(count_20_locations_by_ward$count)))))

colors <- unique_color_df %>% 
  right_join(count_20_locations_by_ward,by='count') %>% 
  select(color)
colors <-  as.vector(as.character(colors[,1]))

leaflet(sf_city_ward) %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = -79.372573, lat = 43.679434, zoom = 10) %>%
  addPolygons(data = count_20_locations_by_ward$geometry, weight = 2,
              popup = paste('Area:',count_20_locations_by_ward$AREA_NAME, "<br>",
                            'Count:',count_20_locations_by_ward$count, "<br>"),
              fillColor = colors) %>%
  addLegend(position = 'bottomright',
            colors=unique_color_df$color,
            labels=unique_color_df$count,
            title='Count for top 20 infractions')
```


### 3.3.2.c Geographic distribution by ward for top 20 infractions (revenue)
### Interpretation: find the revenue of infractions within each ward and find the top 20 wards

```{r}
# First retrieve aggregated sum revenues per location
revenue_df <- parking_ticket[(parking_ticket$location2 %in% top_20_locations_geocoded$location2),] %>%
  group_by(location2) %>%
  summarise(revenue=sum(set_fine_amount)) %>%
  as.data.frame() %>%
  right_join(top_20_locations_geocoded,by='location2')

# Then create a table with summed revenue

revenue_20_locations_by_ward <- revenue_df %>%
  group_by(AREA_NAME) %>%
  summarise(sum=sum(revenue)) %>%
  select(AREA_NAME,sum) %>%
  right_join(sf_city_ward,by='AREA_NAME') %>%
  select(AREA_NAME,sum,geometry) %>%
  replace(is.na(.), 0)

# Finally, plot it out

unique_color_df <- data.frame(sum = sort(unique(revenue_20_locations_by_ward$sum)),
                              color = gsub(".{2}$","",topo.colors(length(unique(revenue_20_locations_by_ward$sum)))))

colors <- unique_color_df %>% 
  right_join(revenue_20_locations_by_ward,by=c('sum')) %>% 
  select(color)
colors <-  as.vector(as.character(colors[,1]))

leaflet(sf_city_ward) %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = -79.372573, lat = 43.679434, zoom = 10) %>%
  addPolygons(data = revenue_20_locations_by_ward$geometry, weight = 2,
              popup = paste('Area:',revenue_20_locations_by_ward$AREA_NAME, "<br>",
                            'Count:',revenue_20_locations_by_ward$sum, "<br>"),
              fillColor = colors) %>%
  addLegend(position = 'bottomright',
            colors=unique_color_df$color,
            labels=unique_color_df$sum,
            title='Revenue for top 20 infractions')
```